Gut-resident microorganisms and their genes are associated with cognition and neuroanatomy in children

### 2023-10-10 manuscript/main.tex
not possible -> challenging

In addition to improved species-level resolution, shotgun metagenomic
sequencing also enables gene-functional insight. We showed that
genes for the metabolism of SCFAs, both their degradation and synthesis,
are associated with cognitive function scores (Figure 2). 
SCFAs are produced by anaerobic fermentation of dietary fiber
and have been linked with immune system regulation as well as directly
with brain function \cite{dalileRoleShortchainFatty2019}.
Microbial metabolism of other potentially neuroactive molecules
were also associated with cognitive ability. For example, we were
particularly interested in genes for metabolizing glutamate, which is a
critical neurotransmitter controlling neuronal excitatory/inhibitory
signaling along with gamma-aminobutyric acid (GABA), and their balance
in the brain controls neural plasticity and learning, particularly in
the developing brain \cite{cohenkadoshLinkingGABAGlutamate2015,palomo-buitragoGlutamateInteractionsObesity2019}.
However, while the
differential abundance of genes for the metabolism of neuroactive
compounds like these is suggestive, it is difficult to reason about the
relationship between levels of these genes and the gut concentrations of
the molecules their product enzymes act on since the presence or absence
of metabolites may select for microbes with the ability to degrade
or produce that metabolite respectively.
For this reason, future studies coupling shotgun
metagenomics with stool metabolomics could improve our understanding of
the relationship between microbial metabolism and cognitive development.
Serum metabolomic analysis may be even more informative,
since the metabolites measured in stool are not necessarily indicative
of exposure, but drawing sufficient blood from children,
especially very young children, is typically not permitted.
Further, strain-level analysis linking specific gene content in species
of interest could further refine targeted efforts at identifying
specific metabolic signatures of microbe-brain interactions.

>>>>

In addition to improved species-level resolution, shotgun metagenomic
sequencing also enables gene-functional insight. We showed that
genes for the metabolism of SCFAs, both their degradation and synthesis,
are associated with cognitive function scores (Figure 2). 
SCFAs are produced by anaerobic fermentation of dietary fiber
and have been linked with immune system regulation as well as directly
with brain function \cite{dalileRoleShortchainFatty2019}.
Microbial metabolism of other potentially neuroactive molecules
were also associated with cognitive ability. For example, we were
particularly interested in genes for metabolizing glutamate, which is a
critical neurotransmitter controlling neuronal excitatory/inhibitory
signaling along with gamma-aminobutyric acid (GABA), and their balance
in the brain controls neural plasticity and learning, particularly in
the developing brain \cite{cohenkadoshLinkingGABAGlutamate2015,palomo-buitragoGlutamateInteractionsObesity2019}.
However, while the
differential abundance of genes for the metabolism of neuroactive
compounds like these is suggestive, it is difficult to reason about the
relationship between levels of these genes and the gut concentrations of
the molecules their product enzymes act on since the presence or absence
of metabolites may select for microbes with the ability to degrade
or produce that metabolite respectively.
For this reason, future studies coupling shotgun
metagenomics with stool metabolomics could improve our understanding of
the relationship between microbial metabolism and cognitive development.
Serum metabolomic analysis may be even more informative,
since the metabolites measured in stool are not necessarily indicative
of exposure, but drawing sufficient blood from children,
especially very young children, is challenging.
Further, strain-level analysis linking specific gene content in species
of interest could further refine targeted efforts at identifying
specific metabolic signatures of microbe-brain interactions.
